The increasing cost of happiness

### 2020-11-23 src/draft0.Rmd
Fixed deflation multiplier and calculated results for all years 2002-2018

Where $X_i$ was an individual's household wealth (\$) as well as other covariates (age, age^2^, gender, education), and $y_i$ was an individual's wellbeing.

@@

The piecewise model was a simple extension of this to include a free parameter to represent the changepoint in wealth ($\omega$) as well as the slope after the change point ($\beta_2$):

$$
\mu_i = \beta_0 + \beta_1 (x_i - \omega_t) (x_i ≤ \omega_t) + \beta_2 (x_i - \omega_t) (x_i > \omega_t)
$$
<br>

@@

```{r read_write_covs}
happywealth_covs <- read_rds("../results/happywealth_covs.rds")
```

@@

happywealth_covs %>%
  filter(!student, (!is.na(losat)|!is.na(gh9))) %>%
  mutate(
    Sex = if_else(male, "Male", "Female"),
    SEIFA = as.integer(extract_numeric(SEIFA)),
    SEIFA = if_else(SEIFA > 0, SEIFA, NA_integer_),
    Education = fct_collapse(
      edu,
      `High school or less` = c("None", "In school", "Highschool", "Trade"),
      Graduate = c("Grad", "Postgrad")
      ),
    Workforce = case_when(
      workforce & !unemployed ~ "Employed",
      unemployed ~ "Unemployed",
      TRUE ~ "Not in labour force"
      ),
    `Relationship status` = case_when(
      single ~ "Single",
      coupled ~ "Married",
      TRUE ~ "Divorced/Widow"
      ),
    Workforce = fct_relevel(Workforce, "Employed", "Unemployed"),
    `Relationship status` = fct_relevel(`Relationship status`, "Single", "Married")
    ) %>%
  select(
    year, 
    # `Life satisfaction` = losat, `Happiness` = gh9, 
    # `Household wealth ($)` = adj_hifdip, 
    Sex, `Age (years)` = age, Education, Workforce, `Relationship status`, 
    `Chronic illness` = chronic, SEIFA, `Household size` = hhsize,
  ) %>% 
  tbl_summary(
    by = year,
    missing = "no",
    statistic = 
      list(all_continuous() ~ "{median} ({p25}-{p75})",
           all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 0,
    missing_text = "(Missing)" 
    ) %>%
  modify_header(stat_by = "{level}\nN = {style_number(n)}") %>%
  as_flex_table() %>%
  set_caption(caption = "Table 1. Demographic characteristics") %>%
  align(part = "body") %>%
  valign(valign = "top", part = "header") %>%
  fontsize(size = 9.5, part = "all") %>%
  width(j = 1, width = 1.5) %>%
  width(j = 2:6, width = 1)
```

@@

p4 <- happywealth_covs %>%
  rename(Happiness = gh9, `Household wealth` = adj_hifdip) %>% 
  filter(!student) %>%
  plot_inflection_custom(Happiness, `Household wealth`, 
                         knots = c(32, 46, 52.5, 86, 124.5)) +
  scale_fill_manual(values = "#C79999") +
  scale_color_manual(values = "#8F2727")
```
```{r figure_1, fig.height=10, fig.width=9}
p1 / p2 / p3 / p4
```

@@

##### Figure 2. Model evidence (mean difference of WAIC~linear~ — WAIC~piecewise~)

```{r plot_waic_losat}
waic_result <- read_rds("../results/waic_losat.RDS")

d1 <- data.frame(fy2018 = rnorm(n = 1e5, 
                                     mean = waic_result[2, 1] * -2, 
                                     sd   = waic_result[2, 2] *  2)/10,
                 fy2014 = rnorm(n = 1e5, 
                                     mean = waic_result[4, 1] * -2, 
                                     sd   = waic_result[4, 2] *  2)/10,
                 fy2010 = rnorm(n = 1e5, 
                                     mean = waic_result[6, 1] * -2, 
                                     sd   = waic_result[6, 2] *  2)/10,
                 fy2006 = rnorm(n = 1e5, 
                                     mean = waic_result[8, 1] * -2, 
                                     sd   = waic_result[8, 2] *  2)/10,
                 fy2002 = rnorm(n = 1e5, 
                                     mean = waic_result[10, 1] * -2, 
                                     sd   = waic_result[10, 2] *  2)/10
                )

color_scheme_set("blue")
p5 <- mcmc_intervals(x = d1,
  pars = colnames(d1),
  prob = 0.0,
  prob_outer = .90
  ) +
  scale_x_continuous(
    limits = c(-10, 40),
    labels = function(x) {x*10}) +
  geom_vline(aes(xintercept = 0), color = "grey90") +
  theme_test() +
  labs(
    subtitle = "Satisfaction",
    x = "")
```

```{r plot_waic_gh9}
waic_result <- read_rds("../results/waic_comparisons_gh9_percentiles.RDS")

d1 <- data.frame(fy2018 = rnorm(n = 1e5, 
                                mean = waic_result[10, 1] * 2, 
                                sd   = waic_result[10, 2] *  2)/10,
                 fy2014 = rnorm(n = 1e5, 
                                mean = waic_result[8, 1] * 2, 
                                sd   = waic_result[8, 2] *  2)/10,
                 fy2010 = rnorm(n = 1e5, 
                                mean = waic_result[6, 1] * 2, 
                                sd   = waic_result[6, 2] *  2)/10,
                 fy2006 = rnorm(n = 1e5, 
                                mean = waic_result[4, 1] * 2, 
                                sd   = waic_result[4, 2] *  2)/10,
                 fy2002 = rnorm(n = 1e5, 
                                mean = waic_result[2, 1] * 2, 
                                sd   = waic_result[2, 2] *  2)/10
                 )

color_scheme_set("red")
p6 <- mcmc_intervals(x = d1,
           pars = colnames(d1),
           prob = 0.00,
           prob_outer = 0.90
           ) +
  scale_x_continuous(
    limits = c(-10, 40),
    labels = function(x) {x*10}) +
  geom_vline(aes(xintercept = 0), color = "grey90") +
  theme_test() +
  labs(
    subtitle = "Happiness",
    x = "WAIC difference (90% interval)")
```

@@

```
Figure 2 legend: Differences in posterior evidence for a linear fit over a piecewise fit (WAIC) for satisfaction (blue) and happiness (red). The filled circle indicates the mean of the distribution and the horizontal bars represents the 90% credible interval. A difference greater than zero is support for the linear model and below zero is support for the piecewise model.
```

@@

##### Figure 3. Posterior distributions of &omega; (90% distribution)  
```{r figure_3, fig.width = 6, fig.height = 4.5, dpi=300}
hifdip_gh9_fixed_fit <- read_rds("../results/hifdip_gh9_fixed_full_covs.RDS")

as.data.frame(hifdip_gh9_fixed_fit) %>%
  select(b_alpha_Intercept, b_alpha_year2006, b_alpha_year2010, 
         b_alpha_year2014, b_alpha_year2018) %>%
  transmute(
    fy2002 = b_alpha_Intercept,
    fy2006 = mean(b_alpha_Intercept) + b_alpha_year2006,
    fy2010 = mean(b_alpha_Intercept) + b_alpha_year2010,
    fy2014 = mean(b_alpha_Intercept) + b_alpha_year2014,
    fy2018 = mean(b_alpha_Intercept) + b_alpha_year2018
  ) %>%
  mutate_all(.funs = ~brms::inv_logit_scaled(., 0, 150)*10) -> posteriors

@@

mcmc_areas_ridges(
  posteriors,
  pars = rev(colnames(posteriors)),
  prob = 0.90, # 80% intervals
  prob_outer = .99, # 99%
  point_est = "mean"
  ) +
  theme_test() +
  geom_vline(aes(xintercept = mean(posteriors$fy2002)), color = "gray", linetype = 3) +
  labs(
    x = "Household wealth ($000s)")
```

@@

```{r, table_2, ft.align="left"}
happywealth_covs %>%
  select(year, xwaveid, adj_hifdip) %>%
  mutate(
    year = paste0("fy", year),
    adj_hifdip = adj_hifdip/1000) %>%
  left_join(
    gather(
      summarise_all(posteriors, list(median)), year, change_point
    )) %>%
  group_by(year) %>%
  summarise(
    `median ($000s)` = round(median(adj_hifdip), 2),
    `mean ($000s)` = round(mean(adj_hifdip), 2),
    `change point ($000s)` = round(mean(change_point), 2),
    `%` = round(mean(adj_hifdip > mean(change_point))*100, 1)
  ) %>%
  flextable() %>%
  set_caption("Table 2. Household wealth, change points and percentile") %>%
  autofit()
```

@@

>>>>

Where $X_i$ was an individual's household wealth (\$) as well as other covariates (age, age^2^, sex, education), and $y_i$ was an individual's wellbeing.

@@

The piecewise model was a simple extension of this to include a free parameter to represent the changepoint in wealth ($\omega$) as well as the slope before the change point ($\beta_1$) and after the change point ($\beta_2$):

$$
\mu_i = \beta_0 + \beta_1 (x_i - \omega) (x_i ≤ \omega) + \beta_2 (x_i - \omega) (x_i > \omega) + \beta_3 X_i
$$

Where $X_i$ were covariates for age, age^2^, sex and education.

@@

```{r read_write_covs}
happywealth_covs <- read_rds("../data/happywealth_covs.rds")
```

@@

happywealth_covs %>%
  filter(year %in% c(2002, 2006, 2010, 2014, 2018)) %>%
  filter(!student, (!is.na(losat)|!is.na(gh9))) %>%
  mutate(
    Sex = if_else(male, "Male", "Female"),
    SEIFA = as.integer(extract_numeric(SEIFA)),
    SEIFA = if_else(SEIFA > 0, SEIFA, NA_integer_),
    Education = fct_collapse(
      edu,
      `High school or less` = c("None", "In school", "Highschool", "Trade"),
      Graduate = c("Grad", "Postgrad")
      ),
    Workforce = case_when(
      workforce & !unemployed ~ "Employed",
      unemployed ~ "Unemployed",
      TRUE ~ "Not in labour force"
      ),
    `Relationship status` = case_when(
      single ~ "Single",
      coupled ~ "Married",
      TRUE ~ "Divorced/Widow"
      ),
    Workforce = fct_relevel(Workforce, "Employed", "Unemployed"),
    `Relationship status` = fct_relevel(`Relationship status`, "Single", "Married")
    ) %>%
  select(
    year, 
    `Life satisfaction` = losat, `Happiness` = gh9,
    # `Household wealth ($)` = adj_hifdip, 
    Sex, `Age (years)` = age, Education, Workforce, `Relationship status`, 
    `Chronic illness` = chronic, SEIFA, `Household size` = hhsize,
  ) %>% 
  tbl_summary(
    by = year,
    missing = "no",
    statistic = 
      list(all_continuous() ~ "{median} ({p25}-{p75})",
           all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 0,
    missing_text = "(Missing)" 
    ) %>%
  modify_header(stat_by = "{level}\nN = {style_number(n)}") %>%
  as_flex_table() %>%
  set_caption(caption = "Table 1. Demographic characteristics") %>%
  align(part = "body") %>%
  valign(valign = "top", part = "header") %>%
  fontsize(size = 9.5, part = "all") %>%
  width(j = 1, width = 1.5) %>%
  width(j = 2:6, width = 1)
```

@@

p4 <- happywealth_covs %>%
  rename(Happiness = gh9, `Household wealth` = adj_hifdip) %>% 
  filter(!student) %>%
  plot_inflection_custom(Happiness, `Household wealth`, 
                         knots = c(43.7, 49.2, 47.4, 53.2, 72.3)) +
  scale_fill_manual(values = "#C79999") +
  scale_color_manual(values = "#8F2727")
```
```{r figure_1, fig.height=10, fig.width=9}
p1 / p2 / p3 / p4
```

@@

##### Figure 2. Linear model evidence (WAIC~linear~ — WAIC~piecewise~)

```{r plot_waic_losat}
waic_result <- read_rds("../results/model_comparisons_waic_R2_losat.RDS")

waic_result %>%
  select(model, year, elpd_diff, se_diff) %>%
  filter(elpd_diff != 0) %>%
  filter(year %in% c(2002, 2006, 2010, 2014, 2018)) %>%
  rowwise() %>%
  mutate(samples = list(rnorm(n = 1e5, 
                              mean = elpd_diff * -2, 
                              sd   = se_diff *  2)),
         year = paste0("AD", year)
         ) %>%
  select(year, samples) %>%
  group_by(year) %>%
  unnest(samples) %>% 
  mutate(sample_id = c(1:1e5)) %>%
  pivot_wider(id_col = "sample_id", names_from = "year", values_from = "samples") %>%
  select(-sample_id) -> d1

# "fix" d1
colnames(d1) <- rev(colnames(d1))

color_scheme_set("blue")

p5 <- mcmc_intervals(x = d1,
  pars = colnames(d1),
  prob = 0.0,
  prob_outer = .95
  ) +
  scale_x_continuous(
    limits = c(-200, 150)
    ) +
  geom_vline(aes(xintercept = 0), color = "grey90") +
  theme_test() +
  labs(
    subtitle = "Satisfaction",
    x = "")
```

```{r plot_waic_gh9}
waic_result <- read_rds("../results/model_comparisons_waic_R2_gh9.RDS")

waic_result %>%
  select(model, year, elpd_diff, se_diff) %>%
  filter(elpd_diff != 0) %>%
  filter(year %in% c(2002, 2006, 2010, 2014, 2018)) %>%
  rowwise() %>%
  mutate(samples = list(rnorm(n = 1e5, 
                              mean = elpd_diff * 2, 
                              sd   = se_diff *  2)),
         year = paste0("AD", year)
         ) %>%
  select(year, samples) %>%
  group_by(year) %>%
  unnest(samples) %>% 
  mutate(sample_id = c(1:1e5)) %>%
  pivot_wider(id_col = "sample_id", names_from = "year", values_from = "samples") %>%
  select(-sample_id) -> d2

p6 <- mcmc_intervals(x = d2,
           pars = rev(colnames(d2)),
           prob = 0.00,
           prob_outer = 0.95
           ) +
  scale_x_continuous(
    limits = c(-200, 150)
    ) +
  geom_vline(aes(xintercept = 0), color = "grey90") +
  theme_test() +
  labs(
    subtitle = "Happiness",
    x = "WAIC difference (95% interval)")
```

@@

```
Figure 2 legend: Differences in posterior evidence for a linear fit over a piecewise fit (WAIC) for satisfaction (blue) and happiness (red). The filled circle indicates the mean of the distribution and the horizontal bars represents the 95% credible interval. A difference greater than zero is support for the linear model and a difference below zero is support for the piecewise model.
```

@@

##### Figure 3. Posterior distributions of &omega; (95% distribution)  
```{r figure_3, fig.width = 6, fig.height = 4.5, dpi=300}
path_to_fits <- list.files(
  path = "../results",#'~/Dropbox/HILDA/data',
  pattern = '^fit_cp_hifdip_gh9_cov_.*.RDS$',
  full.names = TRUE
)

posterior_alphas <- list()
for (pathtofile in path_to_fits) {
  df <- read_rds(pathtofile) %>%
    as.data.frame() %>%
    select(b_alpha_Intercept)

  colnames(df) <- paste0("b_alpha_", substring(pathtofile, 34, 37))

  posterior_alphas <- bind_cols(posterior_alphas, df)
}

posterior_alphas %>%
  mutate_all(.funs = ~brms::inv_logit_scaled(., 0, 100)*10) -> posterior_omegas

# select posteriors for plotting
posteriors_selected <- posterior_omegas %>%
  mutate(b_alpha_2010 = (b_alpha_2010 + b_alpha_2011)/2) %>%
  select(b_alpha_2002, b_alpha_2006, b_alpha_2010, b_alpha_2014, b_alpha_2018)

@@

mcmc_areas_ridges(
  posteriors_selected,
  pars = rev(colnames(posteriors_selected)),
  prob = 0.95, # 80% intervals
  prob_outer = .99 # 99%
  ) +
  theme_test() +
  geom_vline(aes(xintercept = median(posteriors_selected$b_alpha_2002)), 
             color = "gray", 
             linetype = 3) +
  labs(x = "Household wealth ($000s)")
```

@@

```{r, table_2, ft.align="left"}
# Trim posteriors for 2003 outlier
posterior_omegas %>%
  mutate(b_alpha_2003 = (b_alpha_2002 + b_alpha_2004)/2) -> posteriors_trimmed

# Tabulate against median wealth
happywealth_covs %>%
  select(year, xwaveid, adj_hifdip) %>%
  mutate(
    year = paste0("b_alpha_", year),
    adj_hifdip = adj_hifdip/1000) %>%
  left_join(
    gather(
      summarise_all(posteriors_trimmed, list(median)), year, change_point
    )) %>%
  mutate(
    year = extract_numeric(year)
    ) %>%
  group_by(year) %>%
  summarise(
    `Median household wealth ($000s)` = round(median(adj_hifdip), 2),
    # `mean ($000s)` = round(mean(adj_hifdip), 2),
    `Cost of happiness ($000s)` = round(mean(change_point), 2),
    `Percent pop. afford happiness` = round(mean(adj_hifdip > mean(change_point))*100, 1)
    ) -> results

results %>%
  filter(year %in% c(2002, 2006, 2010, 2014, 2018)) %>%
  flextable() %>%
  set_caption("Table 2. Household wealth, change points and percentile") %>%
  autofit()
```

@@

##### Figure 4.  
```{r figure_4}
results %>%
  gather("key", "value", -year) %>%
  mutate(
    key = paste0(key, "\n"),
    key = fct_relevel(key, "Median household wealth ($000s)\n")) %>%
  select(key) %>% distinct() %>%
  mutate(string = c("GFC occurred\nin 2009", "", "")) -> df.note

p <- results %>%
  gather("key", "value", -year) %>%
  mutate(
    key = paste0(key, "\n"),
    key = fct_relevel(key, "Median household wealth ($000s)\n")) %>%
  ggplot(aes(x = year, y = value, color = key)) +
    geom_line(aes(group = key), size = 2, alpha = 0.5) +
    # geom_smooth(aes(group = key), se = F, span = 2) +
    labs(title = "Wealth & happiness in Australia between 2002-2018",
         caption = "Source: HILDA, Melbourne University",
         x = "  ", y = "") +
    geom_vline(aes(xintercept = 2009), color = "red", linetype = 3) +
    geom_text(data = df.note, aes(x = 2009.5, y = 55, label = string),
              colour = "red", size = 3, hjust = "left") +
    facet_wrap(~key)

# p.wsj <- p +
#   ggthemes::scale_colour_wsj("colors6", "") +
#   ggthemes::theme_wsj() +
#   theme(legend.position = "none",
#         plot.title = element_text(size = 19),
#         plot.caption = element_text(size = 12),
#         strip.text = element_text(size = 12))

p.econ <- p + 
  labs(subtitle = "  ") +
  ggthemes::scale_color_economist() +
  ggthemes::theme_economist(base_family="Verdana", base_size = 8) +
  theme(legend.position = "none",
        strip.text = element_text(size = 10)
        )
p.econ
```
